---
title: "BARCS regression for continuous and multivariable CRISPR screens"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BARCS regression for continuous and multivariable CRISPR screens}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

CB2's original `measure_sgrna_stats()` answers a two-group question. BARCS
(Beta-binomial Analysis and Regression for CRISPR Screens) answers a
coefficient question: does guide abundance change with dose, time, an ordered
phenotype, or an adjusted treatment effect?

The regression interface is additive and does not change the original
workflow. It keeps guide count as the response and retains CB2's
beta-binomial treatment of library depth and between-sample heterogeneity.
What changes is the mean model: a standard R formula can contain a continuous
sample phenotype, adjustment variables, interactions, splines, or categorical
contrasts.

This is a statistical extension rather than an assertion that the default
functions return the same number. If the completed original weighted group
proportions are written as \(\widehat p_A\) and \(\widehat p_B\), with
estimated variances \(V_A\) and \(V_B\), then a saturated two-cell
generalized least-squares representation has coefficient vector
\[
(\widehat p_A,\widehat p_B-\widehat p_A)^\top
\]
and slope variance \(V_A+V_B\). Its slope t statistic is therefore exactly the
original CB2 statistic. This is a legacy summary-scale compatibility identity,
not proof that default `bbreg()` strictly nests the old estimator. On the logit
scale, Taylor expansion around
\(p_A=p_B=p\) multiplies both the effect and standard error by
\(\{p(1-p)\}^{-1}\), which cancels. Under common dispersion, local
alternatives, and increasing numbers of independent biological libraries, the
logit statistic is first-order equivalent. Increasing sequencing depth at
fixed replication is not sufficient. In finite samples the estimators differ:
original CB2 uses separate group weighting and Welch-style degrees of freedom;
`bbreg()` estimates one guide-wise dispersion across the entire design and
uses residual sample degrees of freedom.

The model-based covariance treats the estimated guide dispersion as a fixed
plug-in value. Student t degrees of freedom do not formally propagate
dispersion-estimation uncertainty, so small-sample calibration should be
checked with simulations, phenotype permutations, or prespecified controls.

| Feature | Original CB2 | BARCS |
|---|---|---|
| Design | Two group labels | Full-rank R model matrix |
| Effect | Group difference | Conditional log-odds coefficient or contrast |
| Covariates | Pairwise design | Dose, time, batch, donor, interactions |
| Dispersion | Group-specific weighting | One guide-wise rho across the design |
| Inference | Welch-style two-group test | Student t with residual sample df |

## Simulated dose-response screen

```{r}
library(CB2)

set.seed(20260723)
design <- data.frame(
  sample_name = sprintf("sample_%02d", 1:18),
  dose = rep(seq(-1.5, 1.5, length.out = 6), 3),
  batch = factor(rep(letters[1:3], each = 6))
)
totals <- round(runif(nrow(design), 40000, 90000))

simulate_guide <- function(intercept, slope, rho = 0.0015) {
  mu <- plogis(intercept + slope * design$dose)
  precision <- 1 / rho - 1
  p <- rbeta(length(mu), mu * precision, (1 - mu) * precision)
  rbinom(length(mu), totals, p)
}

counts <- rbind(
  geneA_sg1 = simulate_guide(-7.0, 0.9),
  geneA_sg2 = simulate_guide(-6.8, 0.8),
  geneB_sg1 = simulate_guide(-7.1, 0),
  geneB_sg2 = simulate_guide(-6.9, 0)
)
```

Fit one guide with beta-binomial regression:

```{r}
fit <- bbreg(
  count = counts[1, ],
  total = totals,
  formula = ~ scale(dose) + batch,
  data = design
)
summary(fit)
bb_contrast(fit, c("scale(dose)" = 1))
```

Or analyze every guide and apply Benjamini-Hochberg correction:

```{r}
bb_result <- bb_screen(
  counts = counts,
  totals = totals,
  data = design,
  formula = ~ scale(dose) + batch,
  term = "scale(dose)",
  gene = sub("_sg[0-9]+$", "", rownames(counts))
)
bb_result
```

If the library contains at least 20 prespecified negative-control guides, use
them to diagnose and, when necessary, scale an inflated t-statistic tail:

```{r, eval=FALSE}
bb_result <- bb_calibrate_controls(
  bb_result,
  control = guide_annotation$gene == "Non-Targeting Control"
)
attr(bb_result, "control_scale")
```

The function matches the empirical 95th percentile of absolute control t
statistics to the corresponding Student t cutoff. It never reduces the
standard errors by default, and retains the original inferential columns with
a `raw_` prefix. This is a calibration tool, not a substitute for a repeated-
measure or joint multinomial model when libraries are correlated.
All finite control guides must share one residual degree of freedom, which
holds when every guide is fitted with the same complete design. Heterogeneous
guide designs require separate calibration strata or a different procedure.

The coefficient is a log-odds change per predictor unit. Its t reference has
sample-level residual degrees of freedom; read depth does not substitute for
biological replication. For a negative-binomial benchmark, run the official
MAGeCK-MLE tool separately with the same sample design rather than embedding a
partial MAGeCK reimplementation inside CB2.
