---
title: "Continuous and multivariable phenotypes in CB2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Continuous and multivariable phenotypes in CB2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

CB2's original `measure_sgrna_stats()` answers a two-group question. CB2-Reg
answers a coefficient question: does guide abundance change with dose, time,
an ordered phenotype, or an adjusted treatment effect?

The regression interface is additive and does not change the original
workflow. It keeps guide count as the response and retains CB2's
beta-binomial treatment of library depth and between-sample heterogeneity.
What changes is the mean model: a standard R formula can contain a continuous
sample phenotype, adjustment variables, interactions, splines, or categorical
contrasts.

| Feature | Original CB2 | CB2-Reg |
|---|---|---|
| Design | Two group labels | Full-rank R model matrix |
| Effect | Group difference | Conditional log-odds coefficient or contrast |
| Covariates | Pairwise design | Dose, time, batch, donor, interactions |
| Variance | Beta-binomial | Same beta-binomial principle |
| Inference | Two-group test | Student t with sample-level residual df |

## Simulated dose-response screen

```{r}
library(CB2)

set.seed(20260723)
design <- data.frame(
  sample_name = sprintf("sample_%02d", 1:18),
  dose = rep(seq(-1.5, 1.5, length.out = 6), 3),
  batch = factor(rep(letters[1:3], each = 6))
)
totals <- round(runif(nrow(design), 40000, 90000))

simulate_guide <- function(intercept, slope, rho = 0.0015) {
  mu <- plogis(intercept + slope * design$dose)
  precision <- 1 / rho - 1
  p <- rbeta(length(mu), mu * precision, (1 - mu) * precision)
  rbinom(length(mu), totals, p)
}

counts <- rbind(
  geneA_sg1 = simulate_guide(-7.0, 0.9),
  geneA_sg2 = simulate_guide(-6.8, 0.8),
  geneB_sg1 = simulate_guide(-7.1, 0),
  geneB_sg2 = simulate_guide(-6.9, 0)
)
```

Fit one guide with beta-binomial regression:

```{r}
fit <- bbreg(
  count = counts[1, ],
  total = totals,
  formula = ~ scale(dose) + batch,
  data = design
)
summary(fit)
bb_contrast(fit, c("scale(dose)" = 1))
```

Or analyze every guide and apply Benjamini-Hochberg correction:

```{r}
bb_result <- bb_screen(
  counts = counts,
  totals = totals,
  data = design,
  formula = ~ scale(dose) + batch,
  term = "scale(dose)",
  gene = sub("_sg[0-9]+$", "", rownames(counts))
)
bb_result
```

If the library contains at least 20 prespecified negative-control guides, use
them to diagnose and, when necessary, scale an inflated t-statistic tail:

```{r, eval=FALSE}
bb_result <- bb_calibrate_controls(
  bb_result,
  control = guide_annotation$gene == "Non-Targeting Control"
)
attr(bb_result, "control_scale")
```

The function matches the empirical 95th percentile of absolute control t
statistics to the corresponding Student t cutoff. It never reduces the
standard errors by default, and retains the original inferential columns with
a `raw_` prefix. This is a calibration tool, not a substitute for a repeated-
measure or joint multinomial model when libraries are correlated.

The coefficient is a log-odds change per predictor unit. Its t reference has
sample-level residual degrees of freedom; read depth does not substitute for
biological replication. For a negative-binomial benchmark, run the official
MAGeCK-MLE tool separately with the same sample design rather than embedding a
partial MAGeCK reimplementation inside CB2.
